---
title: "Di Zhen (1717719)"
author: "dizhen"
date: "2020.3.28"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Set working directory:

```{r}
setwd("D:/git/DPH112-xjtlu/week05")
```
Data

```{r}
KBI <- c(28,68,59,91,70,38,46,57,89,48,74,78,43,76,72,61,63,77,85,31,79,92,76,91,78,103,99,73,88,64,52,71,41,85,52,68,57,84,91,83,73,57,69,81,71,91,48,94,57,49,88,54,73,87,47,60,65,57,85,28,40,87,80,49,57,32,52,42,49,63,89,67,43,47,70,99,53,78,112,52,68,63,49,42,56,46,72,95,57,88,81,104,88,115,66,92,97,69,112,88)

ADL <- c(39,52,89,57,28,34,42,52,88,90,38,83,30,45,47,90,63,34,76,26,68,85,22,82,80,80,81,30,27,72,46,63,45,77,42,60,33,49,89,72,45,73,58,33,34,90,48,47,32,63,76,79,48,90,55,83,50,44,79,24,40,35,55,45,46,37,47,28,61,35,68,80,43,53,60,63,28,35,37,82,88,52,30,69,52,59,53,65,90,88,66,60,48,82,88,63,79,71,66,81)

MEM <- c(4,33,17,31,35,3,16,6,41,24,22,41,9,33,36,17,14,35,33,13,34,28,12,57,51,20,20,7,27,9,15,52,26,57,10,34,14,30,64,31,24,13,16,17,13,42,7,17,13,32,50,44,57,33,11,24,21,31,30,5,20,15,9,28,19,4,29,23,8,31,65,29,8,14,30,22,9,18,33,25,16,15,16,49,17,38,22,56,12,42,12,21,14,41,24,49,34,38,48,66)

COG <- c(18,9,3,7,19,25,17,26,13,3,13,11,24,14,18,0,16,22,23,18,26,10,16,3,3,18,1,17,27,0,22,13,18,0,19,11,14,15,0,3,19,3,15,21,18,6,23,18,15,15,5,11,9,6,20,11,25,18,20,22,17,27,21,17,17,21,3,21,7,26,6,10,13,18,16,18,27,14,17,13,0,0,18,12,20,17,21,2,0,6,23,7,13,13,14,5,3,17,13,1)

mydata <- data.frame(KBI,ADL,MEM,COG)
str(mydata)
head(mydata)
```

## Question 1

```{r}
# pairs(~KBI + ADL + MEM + COG, data = mydata,
#       lower.panel = NULL,
#       main = "Scatterplot Matrix")

if(!require(car)){install.packages("car")}
library("car")
scatterplotMatrix(~KBI + ADL + MEM + COG, data = mydata,
                  lower.panel = NULL,
      main = "Scatterplot Matrix")
```

## Question 2

KBI is positively correlated with ADL and MEM, slightly negatively correlated with COG. ADL is positively correlated with MEM and negatively correlated with COG. MEM is negatively correlated with COG.

## Question 3

```{r}
mydata.LM1 <- lm(KBI ~ ADL + MEM + COG, data = mydata)
summary(mydata.LM1)
```

The multiple regression equition is:

$KBI = 40.5 + 0.2 \times ADL + 0.6 \times MEM + 0.1 \times COG$

## Question 4

1) The intercept is 40.5. If ADL = MEM = COG = 0 and have their meanings (or in the scope), the KBI is equal to 40.5. If ADL = MEM = COG = 0 is out of scope, the intercept has no practical meaning.

2) The coefficient of ADL is 0.2. This means if the score of total activities of daily living(ADL) increases 1 unit, the expected KBI will increase by 0.2, holding other variables constant.

3) The coefficient of MEM is 0.6. This means if the score of memory and behavioral problems (MEM) increases 1 unit, the expected KBI will increase by 0.6, holding other variables constant.

4) The coefficient of COG is 0.1. This means if the score of cognitive impairment  (COG) increases 1 unit, the expected KBI will increase by 0.1, holding other variables constant.

## Question 5

1) Test the null hypothesis of $H_0: intercept = 0$ at $\alpha = 0.03$:

The output gives the value of the t statistics as 4.0 and the p-value is $1.21 \times 10^{-4}$. The p-value is less than $\alpha = 0.03$ providing us with evidence to reject the null hypothesis.

2) Test the null hypothesis of $H_0: \beta_{ADL} = 0$ at $\alpha = 0.03$:

The output gives the value of the t statistics as 1.9 and the p-value is 0.07. The p-value is more than $\alpha = 0.03$,thus we fail to reject the null hypothesis.

3) Test the null hypothesis of $H_0: \beta_{MEM} = 0$ at $\alpha = 0.03$:

The output gives the value of the t statistics as 4.3 and the p-value is $4.65\times 10^{-5}$. The p-value is less than $\alpha = 0.03$ providing us with evidence to reject the null hypothesis.

4) Test the null hypothesis of $H_0: \beta_{COG} = 0$ at $\alpha = 0.03$:

The output gives the value of the t statistics as 0.4 and the p-value is 0.69. The p-value is more than $\alpha = 0.03$, thus we fail to reject the null hypothesis.

## Question 6

```{r}
confint(mydata.LM1, level = 0.97)
```

1) The 97% confidence interval for intercept is from 18.2 to 62.7. We are 97% confident that the intercept estimate lies between the interval 18.2 and 62.7, because on repeated sampling, 97% of intervals constructed in ths manner will contain the true intercept.

2) The 97% confidence interval for slope of ADL is from 0 to 0.5. We are 97% confident that the slope estimate of ADL lies between the interval 0 and 0.5, because on repeated sampling, 97% of intervals constructed in ths manner will contain the true slope of ADL.

3) The 97% confidence interval for slope of MEM is from 0.3 to 0.8. We are 97% confident that the slope estimate of MEM lies between the interval 0.3 and 0.8, because on repeated sampling, 97% of intervals constructed in ths manner will contain the true slope of MEM.


4) The 97% confidence interval for slope of COG is from -0.5 to 0.8. We are 97% confident that the slope estimate of COG lies between the interval -0.5 and 0.8, because on repeated sampling, 97% of intervals constructed in ths manner will contain the true slope of COG.

## Question 7

```{r}

mydata.NEW1 <- data.frame(ADL = 1, MEM = 2, COG = 3)
mydata.NEW2 <- data.frame(ADL = 40, MEM = 50, COG = 60)
predict(mydata.LM1, mydata.NEW1, 
        interval = "prediction",
        level = 0.97)
predict(mydata.LM1, mydata.NEW1, 
        interval = "confidence",
        level = 0.97)
predict(mydata.LM1, mydata.NEW2, 
        interval = "prediction",
        level = 0.97)
predict(mydata.LM1, mydata.NEW2, 
        interval = "confidence",
        level = 0.97)
```

Suppose there are two new caregivers. One got ADL=1, MEM=2, COG=3, the other got ADL = 40, MEM = 50, COG = 60.

1) For the first caregiver:

The 97% confidence interval for mean KBI is (-0.9, 85.2). The 97% prediction interval for KBI is (21.9, 62.4).

2) For the second caregiver:

The 97% confidence interval for mean KBI is (35.1, 133.2). The 97% prediction interval for KBI is (53.1, 115.1).

## THE END
