---
title: "DiZhen_1717719"
author: "dizhen"
date: "2020/4/18"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Data

```{r}
KBI <- c(28,68,59,91,70,38,46,57,89,48,74,78,43,76,72,61,63,77,85,31,79,92,76,91,78,103,99,73,88,64,52,71,41,85,52,68,57,84,91,83,73,57,69,81,71,91,48,94,57,49,88,54,73,87,47,60,65,57,85,28,40,87,80,49,57,32,52,42,49,63,89,67,43,47,70,99,53,78,112,52,68,63,49,42,56,46,72,95,57,88,81,104,88,115,66,92,97,69,112,88)

ADL <- c(39,52,89,57,28,34,42,52,88,90,38,83,30,45,47,90,63,34,76,26,68,85,22,82,80,80,81,30,27,72,46,63,45,77,42,60,33,49,89,72,45,73,58,33,34,90,48,47,32,63,76,79,48,90,55,83,50,44,79,24,40,35,55,45,46,37,47,28,61,35,68,80,43,53,60,63,28,35,37,82,88,52,30,69,52,59,53,65,90,88,66,60,48,82,88,63,79,71,66,81)

MEM <- c(4,33,17,31,35,3,16,6,41,24,22,41,9,33,36,17,14,35,33,13,34,28,12,57,51,20,20,7,27,9,15,52,26,57,10,34,14,30,64,31,24,13,16,17,13,42,7,17,13,32,50,44,57,33,11,24,21,31,30,5,20,15,9,28,19,4,29,23,8,31,65,29,8,14,30,22,9,18,33,25,16,15,16,49,17,38,22,56,12,42,12,21,14,41,24,49,34,38,48,66)

COG <- c(18,9,3,7,19,25,17,26,13,3,13,11,24,14,18,0,16,22,23,18,26,10,16,3,3,18,1,17,27,0,22,13,18,0,19,11,14,15,0,3,19,3,15,21,18,6,23,18,15,15,5,11,9,6,20,11,25,18,20,22,17,27,21,17,17,21,3,21,7,26,6,10,13,18,16,18,27,14,17,13,0,0,18,12,20,17,21,2,0,6,23,7,13,13,14,5,3,17,13,1)

mydata <- data.frame(KBI,ADL,MEM,COG)
str(mydata)
head(mydata)
```


Visualization

```{r}
pairs(~KBI + ADL + MEM + COG, data = mydata,
      lower.panel = NULL,
      main = "Scatterplot Matrix")
```

## Question 1

```{r}
mydata.LM1 <- lm(KBI ~ ADL + MEM + COG, data = mydata)
summary(mydata.LM1)
```


The multiple regression equition is:

$KBI = 40.5 + 0.2 \times ADL + 0.6 \times MEM + 0.1 \times COG$

The mean KBI when ADL, MEM and COG are zero is 40.5. For every extra unit of ADL, the expected KBI increases by 0.2 points, holding other variables constant. For every extra unit of MEM, the mean KBI increases by 0.6 points, holding other variables constant. For every extra unit of COG, the mean KBI increases by 0.1 points, holding other variables constant.

## Question 2

```{r}
mydata$FITTED <- predict(mydata.LM1, type = "response")
mydata$RESID <- resid(mydata.LM1)
head(mydata)
```

## Question 3

1. Plot the residuals against the fitted values (RvF plot)

```{r}
plot(RESID~FITTED, data = mydata,
     main = "RvF Plot",
     ylab = "Residuals")
```

2. Plot the residuals against each of the predictors


```{r}
pairs(~RESID + ADL + MEM + COG, data=mydata,
      lower.panel=NULL,
      main = "Scatterplot Matrix")
```


In the RvF plot, we do not notice any systematic non-linear deviation.
In the scatterplot matrix, we are only interested in the first row. Neither shows any systematic non-linear deviation.

Conclusion: overall, the assumption of normality is not violated.

## Question 4

Produce a quantile-quantile plot of the residuals.

```{r}
qqnorm(mydata$RESID,
       main = "Normal Quantile-Quantile Plot of Ordianry Residuals")
qqline(mydata$RESID)
```


The graph shows some deviation in the right tail. Although the points hug the qq line quite closely, this deviation affects more than 10 observations. 

Conclusion: The assumption of normality is violated.

## Question 5

1. Plot the residuals against the fitted values (RvF plot)

```{r}
plot(RESID~FITTED, data = mydata,
     main = "RvF Plot",
     ylab = "Residuals")
```

2. Plot the residuals against each of the predictors

```{r}
pairs(~RESID + ADL + MEM + COG, data=mydata,
      lower.panel=NULL,
      main = "Scatterplot Matrix")
```

There is no wedge-shaped patterns or trumpteting.

Conclusion: The assumption of equal variance of residuals is not violated.

## Question 6

Evidence about independence is available from the design of data.

## Question 7
```{r}
pairs(~ADL + MEM + COG, data=mydata,
      lower.panel=NULL,
      main = "Collinearity between Predictors")
```

From the plot, we observe little linear relatioship between predictors. There is negative relationships between ADL and COG, and between MEM and COG. There is positive relationship between MEM and ADL.

```{r}
cor(mydata$ADL,mydata$MEM)
cor(mydata$ADL,mydata$COG)
cor(mydata$MEM,mydata$COG)
```

The correlation between ADL and MEM is 45.57%. The correlation between ADL and COG is -64.70%. The correlation between MEM and COG is -42.98%.

```{r}
# if(!require(car)){install.packages("car")}
library(car)
vif(mydata.LM1)
```
The VIF(ADL), VIF(MEM), VIF(COG) are all < 10, indicating the absence of collienarity.

Conclusion: there is no multicollinearity between the predictors.

## THE END